Magnetic Inhibition of Convection and the Fundamental Properties of Low--Mass Stars

### 2016-02-18 paper/ms.tex
interpolation curve used to compute mass tracks
analysis of observational errors and the knee feature

\usepackage{hyperref}
\hypersetup{
    bookmarks=true,                                  % show bookmarks bar in pdf reader
    pdftitle={A Consistent 10 Myr Age for the Upper Scorpius OB Association},          % set pdf title          
    pdfauthor={Gregory A. Feiden},   % set pdf author
    colorlinks=true,                                 % false = box link, true = colored links
    linkcolor=blue,                                  % color of internal links
    citecolor=blue,                                  % color of citations
    urlcolor=blue                                    % external url color
}

@@

\begin{table}[t]
    \centering
    \caption{Equipartition magnetic field strength predictions at 10 Myr.}
    \begin{tabular*}{0.6\linewidth}{@{\extracolsep{\fill}}c c c}
        \hline\hline\noalign{\smallskip}
        Mass (\msun) & \teff (K) & $\langle Bf \rangle_{\rm eq,\,surf}$ (kG) \\
        \noalign{\smallskip}\hline\noalign{\smallskip}
        0.1  & 2920 & 2.58 \\
        0.2  & 3127 & 2.54 \\
        0.3  & 3260 & 2.56 \\
        0.4  & 3371 & 2.57 \\
        0.5  & 3463 & 2.56 \\
        0.6  & 3551 & 2.52 \\
        0.7  & 3640 & 2.47 \\
        0.8  & 3735 & 2.40 \\
        0.9  & 3836 & 2.29 \\
        1.0  & 3962 & 2.12 \\
        1.1  & 4099 & 1.94 \\
        1.2  & 4340 & 1.80 \\
        1.3  & 4617 & 1.69 \\
        1.4  & 4976 & 1.56 \\
        1.5  & 5565 & 1.29 \\
        1.6  & 6281 & 0.94 \\
        1.7  & 7004 & 0.77 \\
        \noalign{\smallskip}\hline
    \end{tabular*}
    \label{tab:equip_values}
\end{table}

% support with evidence from the literature? perhaps in discussion section?
The interior magnetic field strength profile requires a surface boundary condition: the average surface magnetic field strength $\langle{\rm B}f\rangle$. In previous studies, a range of values were explored and values that reproduce a given stellar fundamental property were eventually selected. 
Here, a different approach is used. Instead of adjusting models to find suitable magnetic field strengths, equipartition estimates of the maximum allowed average surface magnetic field strengths are adopted. The aim is to test the predictive abilities of magnetic stellar evolution models. Equipartition between the surface magnetic field pressure and the local gas pressure is assumed, meaning
\begin{equation}
    B_{\rm eq,\, surf} = \left(8\pi P_{\rm gas,\, surf}\right)^{1/2}.
\end{equation}
Surface gas pressures are defined using the same model atmosphere structures used to specify surface boundary conditions. The ``surface'' is defined as the optical photosphere where $\tau_{5000} = 1$. Values of the surface equipartition magnetic field strengths are listed in Table~\ref{tab:equip_values} for stars at an age of 10 Myr. Since surface gas pressures are dependent on \logg\ and \teff, equipartition magnetic field strengths evolve toward greater values for stars undergoing quasi-hydrostatic contraction. This assumption breaks down for stars with masses above 1.55 $M_{\odot}$, whose outer convection zones disappear within {\bf about} 10 Myr.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=0.85\linewidth]{USco_median_HRD.eps}
    \caption{Hertzsprung-Russell (HR) diagram for Upper Scorpius. A-type through M-type members of Upper Scorpius \citep{Preibisch1999,Preibisch2002,Pecaut2012} are shown as gray points. A median empirical locus for the population is shown as black solid line with the 99\% confidence interval for the median value shown as a light-gray shaded region. Errorbars shown for the majority of stars with \teff\ $< 5\,500$~K are prescribed as described in Sect.~\ref{sec:data}.}
    \label{fig:empirical}
\end{figure}

@@

\section{Results}
\label{sec:results}
An HR diagram for Upper Scorpius is shown in Fig.~\ref{fig:ages}a. Empirical data is represented by the median stellar locus ({\bf light-gray shaded region}) and against three standard stellar evolution isochrones with ages of 5, 10, and 15 Myr ({\bf black dashed lines}). Focusing on {\bf temperature regimes corresponding} A-, F-, and G-type stars {\bf ($5\,750 <$ \teff/K $< 10\,000$)}, standard stellar isochrones predict ages between 5 and 15 Myr, with a median age estimate of approximately 10 Myr. %A median age of roughly 10 Myr is confirmed by comparison of the median stellar locus with stellar model isochrones.

@@

Introducing a magnetic perturbation as described in Sect.~\ref{sec:models}, a 10 Myr isochrone is computed and overlaid in the HR and mass-radius diagrams in Figs.~\ref{fig:ages} and \ref{fig:mrd} (thick yellow solid line). An age of 10 Myr was chosen as a starting point based on the approximate median age of the A-, F-, and G-type stars determined from standard stellar evolution models.

\begin{figure}[!b]
    \centering
    \includegraphics[width=0.90\linewidth]{USCO5_MR.eps}
    \caption{Mass-radius diagram for stars in eclipsing systems. {\bf Black points are from \citet{Kraus2015} and \citet{Alonso2015}, while gray points are taken from \citet{Lodieu2015} and \citet{David2015b}.} Model predictions for magnetic and non-magnetic isochrones are the same as in Fig.~\ref{fig:ages}.}
    \label{fig:mrd}
\end{figure}

Figures~\ref{fig:ages}a and \ref{fig:ages}c (HR diagrams) {\bf show that a 10 Myr isochrone computed with equipartition magnetic fields} is shifted toward cooler temperatures and higher luminosities compared to a 10 Myr non-magnetic isochrone. Inhibition of convection by magnetic fields {\bf cools the stellar surface temperature thereby slowing} the contraction rate of young stars, leading to stars having a larger radius and a larger luminosity at a given age. The combination of cooler surface temperatures and higher luminosities {\bf makes} a 10 Myr magnetic isochrone look nearly identical to a 5 Myr non-magnetic isochrone for effective temperatures {\bf below about 5\,000~K. The magnetic stellar model isochrone lies on top of the 5 Myr standard model isochrone in Figs.~\ref{fig:ages}a and \ref{fig:ages}c. It also lies entirely within} the empirical stellar locus for low-mass stars in Upper Scorpius.

{\bf The magnetic model isochrone converges with the 10 Myr standard model isochrone at warmer effective temperatures reproducing, to some degree, an observed transition in the HR diagram from a 5 Myr to a 10 Myr standard model isochrone. For \teff\ $< 4\,500$~K, the magnetic isochrone closely matches predictions from a 5 Myr standard model isochrone. Compare this to the small segment of the magnetic isochrone above \teff\ $\sim 6\,000$~K, where the magnetic isochrone closely follows the 10 Myr standard model isochrone. Although the model predicted transition appears to occur over approximately the correct effective temperature domain, Fig.~\ref{fig:ages}a shows a sharper transition from a magnetic to a non-magnetic sequence between $4\,500 <$ \teff/K $< 6\,000$ compared to model predictions. This can be more clearly seen in Fig.~\ref{fig:empirical}, where the empirical median stellar locus exhibits a knee around \teff $\sim 4\,500$~K that roughly corresponds to the beginning of the transition region in Fig.~\ref{fig:ages}a. However, a distinct knee feature is absent from the magnetic model isochrone.

In magnetic model space, this temperature domain corresponds to masses between $1.40$ and $1.56$ \msun. Inspection of Fig.~\ref{fig:mrd} reveals this mass range corresponds precisely to where the rapid thinning of the convection zone occurs, as expected.}

Thus, a 10 Myr magnetic isochrone naturally explains the factor of two age difference observed between high- and low-mass stars, provided the high mass stars are not appreciably affected by non-standard physics (e.g., magnetic fields or rotation).

Furthermore, the mass-radius diagram in Figure~\ref{fig:mrd} shows a 10 Myr magnetic isochrone provides reasonable agreement with the two low-mass eclipsing binary systems. It is apparent that magnetic inhibition of convection leads to more significant radius ``inflation'' at higher masses as compared to lower masses. This effect leads to a change in slope of the model predicted mass-radius relationship that is consistent with observations. Although agreement is good, it is not perfect. {\bf Depending on whether mass and radius estimates are adopted from \citet{Kraus2015} or \citet{David2015b}, at 10 Myr the radii of the primary and secondary of UScoCTIO5 are either 3.2\% and 2.1\% \citeauthor{Kraus2015} larger than model predictions or 5.7\% and 6.6\% larger \citeauthor{David2015b}.} Similarly, models at 10 Myr predict radii that are 1.4\% too large and 1.5\% too small for the Ba and Bb components of HD 144548 at 10 Myr, respectively. An age between 9.0 and 9.5 Myr is preferred for UScoCTIO5, while 10 Myr is the best fit age for HD 144548 Ba/Bb.

@@

\section{Discussion}
\label{sec:disc}
An age of {\bf 9 -- 10} Myr gives broad agreement across the HR and mass-radius diagrams. No stellar population in Upper Scorpius {\bf studied here robustly} supports the possibility that the median age of the association is 5 Myr old. As noted above, the median age for F-type stars appears to be a few million years older than stars of other spectral types. This {\bf discrepancy} is {\bf perplexing} and should be explored further, especially considering that potential missing physics in the models would likely conspire to make F-stars older, not younger. Curiously, most F stars in Upper Scorpius are at a stage of evolution where radii increase and subsequently decrease over a period of roughly 4 Myr {\bf (e.g., HD 144548 A). The timescale for this process is} similar to the observed age discrepancy (see Fig.\ \ref{fig:bump} and text below for more details) and may be indicative of potential shortcomings in model physics.

@@

Given that magnetic inhibition of convection is required to explain the anomalously young age of low-mass K- and M-type stars, it is worth considering at what mass or effective temperature the effects related to magnetism can be identified.
%begins to play a significant role in governing stellar structure. 
Looking at Fig.\ \ref{fig:ages}a, early- to mid-G stars follow a 10 Myr standard stellar model isochrone. Instead of continuing the trend, late-G stars appear to deviate toward a 5 Myr isochrone around \teff\ $\sim 5\,750$ K. %, evidenced by a flattening of the empirical locus. 
This may be an artifact of how the empirical locus was computed. However, it may indicate {\bf a transition} where the influence of magnetic inhibition of convection becomes identifiable.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.85\linewidth]{radius_bump.eps}
    \caption{Brief period of radius inflation described in Sect.~\ref{sec:radius} for stars with masses of 1.60\msun\ (blue solid), 1.55\msun\ (green short-dashed), 1.50\msun\ (blue dash-dotted), 1.45\msun\ (yellow long-dashed). Mass tracks are computed without a magnetic field (i.e., non-magnetic configuration). Effective temperatures at these masses are characteristic of F stars at 10 Myr.}
    \label{fig:bump}
\end{figure}

@@

@@

\bibliography{papers}

>>>>

\usepackage{hyperref}
\hypersetup{
	bookmarks=true,
    pdftitle={A Consistent 10 Myr Age for the Upper Scorpius OB Association},          % set pdf title          
    pdfauthor={Gregory A. Feiden},   % set pdf author
    colorlinks=true,                                 % false = box link, true = colored links
    linkcolor=blue,                                  % color of internal links
    citecolor=blue,                                  % color of citations
    urlcolor=blue                                    % external url color
}

@@

\begin{table}[t]
    \centering
    \caption{Equipartition magnetic field strength predictions at 10 Myr.}
    \begin{tabular*}{0.7\linewidth}{@{\extracolsep{\fill}}c c c c}
        \hline\hline\noalign{\smallskip}
        Mass    & \teff & \logg  & $\langle Bf \rangle_{\rm eq,\,surf}$ \\
        \noalign{\smallskip}
        (\msun) &   (K) & (dex)  &   (kG) \\
        \noalign{\smallskip}\hline\noalign{\smallskip}
        0.1 &   3060 &   4.16 &   2.64 \\
  		0.2 &   3261 &   4.19 &   2.51 \\
  		0.3 &   3396 &   4.20 &   2.44 \\
  		0.4 &   3517 &   4.22 &   2.41 \\
  		0.5 &   3639 &   4.24 &   2.38 \\
  		0.6 &   3760 &   4.26 &   2.34 \\
  		0.7 &   3888 &   4.28 &   2.29 \\
  		0.8 &   4031 &   4.29 &   2.22 \\
  		0.9 &   4195 &   4.30 &   2.14 \\
  		1.0 &   4397 &   4.30 &   2.04 \\
  		1.1 &   4641 &   4.30 &   1.92 \\
  		1.2 &   4910 &   4.28 &   1.83 \\
  		1.3 &   5214 &   4.25 &   1.73 \\
  		1.4 &   5569 &   4.19 &   1.58 \\
  		1.5 &   5995 &   4.06 &   1.29 \\
  		1.6 &   6618 &   3.96 &   0.94 \\
  		1.7 &   7403 &   4.15 &   0.76 \\
        \noalign{\smallskip}\hline
    \end{tabular*}
    \label{tab:equip_values}
\end{table}

% support with evidence from the literature? perhaps in discussion section?
The interior magnetic field strength profile requires a surface boundary condition: the average surface magnetic field strength $\langle{\rm B}f\rangle$. In previous studies, a range of values were explored and values that reproduce a given stellar fundamental property were eventually selected. 
Here, a different approach is used. Instead of adjusting models to find suitable magnetic field strengths, equipartition estimates of the maximum allowed average surface magnetic field strengths are adopted. The aim is to test the predictive abilities of magnetic stellar evolution models. Equipartition between the surface magnetic field pressure and the local gas pressure is assumed, meaning
\begin{equation}
    B_{\rm eq,\, surf} = \left(8\pi P_{{\rm gas},\, \tau = 1}\right)^{1/2}.
\end{equation}
Surface gas pressures are defined using the same model atmosphere structures used to specify surface boundary conditions. The ``surface'' is defined as the optical photosphere where $\tau_{5000} = 1$. Values of the surface equipartition magnetic field strengths are listed in Table~\ref{tab:equip_values} for stars at an age of 10 Myr. Since surface gas pressures are dependent on \logg\ and \teff, equipartition magnetic field strengths evolve toward greater values for stars undergoing quasi-hydrostatic contraction. This assumption breaks down for stars with masses above 1.55 $M_{\odot}$, whose outer convection zones disappear within {\bf about} 10 Myr.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=0.85\linewidth]{USco_median_HRD.eps}
    \caption{Hertzsprung-Russell (HR) diagram for Upper Scorpius. A-type through M-type members of Upper Scorpius \citep{Preibisch1999,Preibisch2002,Pecaut2012} are shown as gray points. A median empirical locus for the population is shown as black solid line with the 99\% confidence interval for the median value shown as a light-gray shaded region. Errorbars shown for the majority of stars with \teff~$< 5\,500$~K are prescribed as described in Sect.~\ref{sec:data}.}
    \label{fig:empirical}
\end{figure}

@@

\section{Results}
\label{sec:results}

\subsection{Standard Models}
An HR diagram for Upper Scorpius is shown in Fig.~\ref{fig:ages}a. Empirical data is represented by the median stellar locus ({\bf light-gray shaded region}) and against three standard stellar evolution isochrones with ages of 5, 10, and 15 Myr ({\bf black dashed lines}). Focusing on {\bf temperature regimes corresponding} A-, F-, and G-type stars {\bf ($5\,750 <$ \teff/K $< 10\,000$)}, standard stellar isochrones predict ages between 5 and 15 Myr, with a median age estimate of approximately 10 Myr. %A median age of roughly 10 Myr is confirmed by comparison of the median stellar locus with stellar model isochrones.

@@

\subsection{Magnetic Models}
Introducing a magnetic perturbation as described in Sect.~\ref{sec:models}, a 10 Myr isochrone is computed and overlaid in the HR and mass-radius diagrams in Figs.~\ref{fig:ages} and \ref{fig:mrd} (thick yellow solid line). An age of 10 Myr was chosen as a starting point based on the approximate median age of the A-, F-, and G-type stars determined from standard stellar evolution models.

A 10 Myr magnetic isochrone naturally explains the factor of two age difference observed between high- and low-mass stars, provided the high-mass stars are not appreciably affected by additional non-standard physics (e.g., rotation). 
Figures~\ref{fig:ages}a and \ref{fig:ages}c (HR diagrams) {\bf show that a 10 Myr isochrone computed with equipartition magnetic fields} is shifted toward cooler temperatures and higher luminosities compared to a 10 Myr non-magnetic isochrone. Inhibition of convection by magnetic fields {\bf cools the stellar surface temperature thereby slowing} the contraction rate of young stars. Stars have a larger radius and a higher luminosity at a given age, as a result. The combination of cooler surface temperatures and higher luminosities {\bf makes} a 10 Myr magnetic isochrone look nearly identical to a 5 Myr non-magnetic isochrone for {\bf stars with} effective temperatures {\bf below about 5\,000~K. The magnetic stellar model isochrone lies on top of the 5 Myr standard model isochrone in Figs.~\ref{fig:ages}a and \ref{fig:ages}c and entirely within} the empirical stellar locus for low-mass stars in Upper Scorpius.

{\bf The magnetic model isochrone also correctly converges toward the 10 Myr standard model isochrone at warmer effective temperatures. To some degree, this  convergence reproduces an observed transition in the empirical HR diagram where the age inferred from standard model isochrones shifts from 5 Myr to 10 Myr. Below \teff~$\sim 5\,000$~K, the magnetic isochrone closely matches predictions from a 5 Myr standard model isochrone. Compare this to the small segment of the magnetic isochrone above \teff\ $\sim 6\,000$~K. Here, the magnetic isochrone traces the 10 Myr standard model isochrone. Although the model predicted transition appears to occur over approximately the correct effective temperature domain, Fig.~\ref{fig:ages}a shows a sharper transition from a magnetic to a non-magnetic sequence between $4\,500 <$ \teff/K $< 6\,000$ compared to model predictions. This is more clearly seen in Fig.~\ref{fig:empirical}, where the empirical median stellar locus exhibits a knee around \teff $\sim 4\,500$~K that roughly corresponds to the beginning of the transition region in Fig.~\ref{fig:ages}a. A distinct knee feature is absent from the magnetic model isochrone.}

%In magnetic model space, this temperature domain corresponds to masses between $1.40$ and $1.56$ \msun. Inspection of Fig.~\ref{fig:mrd} reveals this mass range corresponds precisely to where the rapid thinning of the convection zone occurs, as expected.}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.90\linewidth]{USCO5_MR.eps}
    \caption{Mass-radius diagram for stars in eclipsing systems. {\bf Black points are from \citet{Kraus2015} and \citet{Alonso2015}, while gray points are taken from \citet{Lodieu2015} and \citet{David2015b}.} Model predictions for magnetic and non-magnetic isochrones are the same as in Fig.~\ref{fig:ages}.}
    \label{fig:mrd}
\end{figure}

Moreover, inspection of the mass-radius diagram in Figure~\ref{fig:mrd} shows a 10 Myr magnetic isochrone provides reasonable agreement with the two low-mass eclipsing binary systems. It is apparent that magnetic inhibition of convection leads to more significant radius ``inflation'' at higher masses compared to at lower masses. This effect leads to a change in slope of the model predicted mass-radius relationship that is consistent with observations. Although agreement is good, it is not perfect. {\bf Depending on whether mass and radius estimates are adopted from \citet{Kraus2015} or \citet{David2015b}, at 10 Myr the radii of the primary and secondary in UScoCTIO5 are either 3.2\% and 2.1\% (\citeauthor{Kraus2015}) larger than model predictions or 5.7\% and 6.6\% larger (\citeauthor{David2015b}).} Similarly, models at 10 Myr predict radii that are 1.4\% too large and 1.5\% too small for the Ba and Bb components of HD 144548 at 10 Myr, respectively. An age between 9.0 and 9.5 Myr is preferred for UScoCTIO5, while 10 Myr is the best fit age for HD 144548 Ba/Bb.

@@

\section{Discussion}
\label{sec:disc}
An age of {\bf 9 -- 10} Myr gives broad agreement across the HR and mass-radius diagrams. No stellar population in Upper Scorpius {\bf studied here robustly} supports the possibility that the median age of the association is 5 Myr. There are, however, still age discrepancies that must be addressed. As noted above, the median age for F-type stars appears to be a few million years older than stars of other spectral types. This {\bf discrepancy} is {\bf perplexing and deserves further investigation}, especially considering that potential missing physics in the models would likely conspire to make F-stars older, not younger. It should be noted that most F-type stars in Upper Scorpius are at a stage of evolution where models predict radii increase and subsequently decrease over a period of roughly 4 -- 8 Myr. {\bf The timescale for this process is} similar to the observed age discrepancy (see Fig.~\ref{fig:bump} in Sect.~\ref{sec:radius}) and may be indicative of potential shortcomings in existing model physics such as radiative opacities or nuclear reaction cross-sections.

@@

Given that magnetic inhibition of convection is required to explain the anomalously young age of low-mass K- and M-type stars, it is worth considering at which mass or effective temperature the effects related to magnetism cease to play an important role in governing stellar structure. 
Looking at Fig.\ \ref{fig:ages}a, early- to mid-G stars follow a 10 Myr standard stellar model isochrone. Instead of continuing the trend, late-G stars appear to slope toward a 5 Myr isochrone around \teff\ $\sim 5\,750$ K. The empirical sequence exhibits a shallower slope than the model isochrone at 10 Myr.
 %, evidenced by a flattening of the empirical locus.

The shallower slope may be an artifact of how the empirical locus was computed, {\bf but this seems unlikely to be a dominant factor as the HR diagram is well-populated at this effective temperature. It is somewhat under-populated in the immediate vicinity of \teff~$= 5\,500$~K, but is once again well-populated near 4\,500~K. The general trend should therefore be robust against errors in how the empirical locus is computed.}

{\bf Observational errors may also provide a plausible explanation for the shallower slope of the HR diagram in the vicinity of 5\,000~K. Below this temperature, stellar atmospheric opacity becomes dominated by molecular species, particularly TiO in the optical and H$_2$O in the infrared. It can be argued that the effective temperature scale used to determine the properties of K- and M-dwarfs is too cool by about 200 K. \citet{Preibisch1999} constructed an spectral type to effective temperature transformation at a gravity intermediate between dwarf and giant scales using the transformation derived for luminosity class IV objects by \citet{deJager1987}. Comparing their results to the 5 -- 30 Myr pre-main-sequence star temperature scale proposed by \citet{Pecaut2013}, one finds that the temperature scale adopted by \citeauthor{Preibisch1999} is about 100~K cooler for early K-type stars. However, this trend reverses around spectral type M0, with \citeauthor{Preibisch1999} finding temperatures about 100~K warmer than \citet{Pecaut2013} for early-to-late M-type stars. Considering these differences, the sharp knee feature in the HR diagram may be accentuated by the adoption of an erroneous temperature scale, but the general trend that late-K to M-type stars are significantly cooler than theoretical predictions at 10 Myr remains. Furthermore, \citet{Herczeg2015} independently find a temperature scale consistent with \citeauthor{Preibisch1999}'s using PHOENIX BT-Settl models \citep{Allard2011} to infer effective temperatures from infrared photometric colors. The evidence suggests the adopted temperature scale is not the dominant reason for the observed transition in the HR diagram.}

Instead of being the result of observational errors, the shallower slope in the HR diagram may indicate {\bf a transition} where the influence of magnetic inhibition of convection becomes {\bf a critical component governing a star's structure}.

@@

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=0.85\linewidth]{radius_bump.eps}
    \caption{Brief period of radius inflation described in Sect.~\ref{sec:radius} for stars with masses of 1.60\msun\ (blue solid), 1.55\msun\ (green short-dashed), 1.50\msun\ (blue dash-dotted), 1.45\msun\ (yellow long-dashed). Mass tracks are computed without a magnetic field (i.e., non-magnetic configuration). Effective temperatures at these masses are characteristic of F stars at 10 Myr.}
    \label{fig:bump}
\end{figure}

@@

\bibliography{papers.bib}
